Fine scale determinants of wood and leaf strategies in a low-elevation rainforest of India: vertical stratification reflects guild-dependent dynamics

### 2016-08-12 inst/doc/draft.Rmd
v0.1.9035 Internship end

Understanding the determinants of species occurrence and co-occurrence is a long standing issue in ecology [@diamond_ecology_1975]. Tropical rainforests have especially fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Many theories have been proposed to explain how species coexistence can be maintained in ecological communities . Niche theory states that the niche differences of species determine wether they can establish and coexist in communities [@weiher_assembly_1995; @lortie_rethinking_2004], while neutral theory shows that species coexistence is possible even when species are functionaly equivalent, under the sole influence of stochasticity in life, death, reproduction and dispersal [@hubbell_unified_2001]. Both theory can provide realistic expectations of taxonomic diversity [@chave_comparing_2002], and some recent approaches have advocated investigating patterns of functional [@westoby_land-plant_2006; @mcgill_rebuilding_2006] and phylogenetic diversity [@mouquet_ecophylogenetics:_2012] to trace the signature of ecological processes related to specific biological attributes and adaptations. 
These approaches have been applied to resolve the conundrum of high tree diversity in tropical forests [@kraft_functional_2010; @baraloto_using_2012; @hardy_phylogenetic_2012], but studies on these ecosystems still remain scarce. We here investigate taxonomic, functional and phylogenetic diversity in a low elevation rainforest of the Western Ghats biodiversity hotspot [@myers_biodiversity_2000], where no study has addressed the relative influence of neutral and niche-based dynamics on local forest composition.

The variation of functional traits among species reflect different abilities to grow, survive and reproduce in a given environment [@violle_let_2007]. These variation thus reflect fundamental trade-offs in ecological strategies along environmental gradients [@mark_westoby_plant_2002]. For instance, specific leaf area varies along a soil fertility-shade index and represent a trade-off between resource acquisition by photosynthesis and leaf investment in defense and leaf lifespan [@hodgson_is_2011]. Due to common ecological strategies, functional traits will covary along basic leaf [@wright_worldwide_2004] and wood [@chave_towards_2009] economic spectra. The leaf economic spectrum opposes productive ecological strategy to conservative persistent strategy. The wood economic spectrum distinguish species strategies over three axes spanning the basic CSR strategies of @grime_evidence_1977 : (i) competitive species, wiht high conductive efficiency, (ii) stress tolerant species, resistant to embolism, and (iii) disturbance tolerant species, with high mechanical strength. @baraloto_decoupled_2010 highlighted decoupled wood and leaf strategies allowing independant variation of these strategies, but @reich_world-wide_2014 conversely advocated that the two spectra have to integrated in a comprehensive whole plant economic spectrum [@freschet_evidence_2010] opposing fast growing species to more conservative ones. Here we address how varying wood and leaf ecological strategies determine species community assembly in an Indian rainforest.

A major challenge in addressing community assembly is to relate local coexistence to a broader spatial context of species dynamics [@ricklefs_disintegration_2008; @munoz_neutral_2016]. Niche-based dynamics can entail community variation along broad environmental gradients, while local community dynamics can be neutral [@leibold_coexistence_2006]. Local community dynamics can further be influenced by niche differences under the influence of competition [@macarthur_limiting_1967]. In this perspective, @messier_trait_2016 have underlined that the significance of the leaf economic spectrum is basically scale-dependent, which can explain conflicting results across studies performed at different scales. In the same logic, a spatial hierarchy of ecological dynamics can entail a hierarchy of variation in functional traits from broad variation among species along large-scale environmental gradients, to intraspecific response to micro-environmental conditions at very fine scale [@albert_when_2011]. Therefore a multiscale framework is needed to represent the hierarchy of ecological processes from regional to local scale [@lortie_rethinking_2004]. Determining the extent of neutral and niche-based dynamics then implies (i) decomposing diversity patterns within and among communities [@ackerly_trait-based_2007], and (ii) designing appropriate statistical tests to disentangle the influence of local and larger-scale dynamics [@bernard-verdier_community_2012]. We will decompose diversity patterns in a large rainforest plot into sources of variation within and between communities [@kraft_disentangling_2011], and investigate the fine-scale and plot-scale dynamics using appropriate null models. For instance, @messier_trait_2016 argued the scale to which traits are covarying in ecological strategy, more especially they question the existence of the leaf economic spectrum at community scale.

Apart from scale-dependent dynamics in space whithin and among communities, ecological guilds can coexist and occupy distinct niches within communities, while the species within guilds are ecologically similar [@wilson_guilds_1999; @vergnon_niches_2009]. The neutral theory has been basically applied in rainforests to adress diversity patterns in the guild of canopy trees [@hubbell_unified_2001], thus ignoring dynamics of understorey trees. However basic ecological constraints drive resource use strategies and biotic interactions in this ecosystems and imply considering functional strategies and relationship among guilds as well. Low elevation rainforests in South India are remarkably heterogeneous in vertical structure and encompass distinctive guilds across the vertical strata [@pascal_wet_1988]. It thus provide an ideal context to analyze guild-dependent dynamics and ecological strategies in tropical rainforests. We will thus address diversity patterns and community assembly at guild-level.

In this paper, we used a trait-based approach to adress guild and community dynamics within and among local tree assemblages in a tropical rain forest from Western Ghats, India. Although of wet evergreen type, the forest undergoes a long dry season and an intense rainy season under the influence of monsoon. In this context, we expect water stress and competition for resource to be key drivers of forest dynamics. We thus adressed two main questions:

1. Do variations and covariations in functional traits reflect distinct ecological strategies within and among species ?
    i. Do and how functional traits vary among species ?
    ii. Do covariations in functional traits represent axes of ecological strategy within species ?
2. Does functional composition reflect the influence of local abiotic and biotic drivers of community dynamics ?
    i. Do local functional composition ($\alpha$ scale) reflect the influence of abiotic and biotic drivers of community dynamics ?
    ii. Do functional composition variation between communities ($\beta$ scale) reflect the changing of abiotic and biotic conditions ?

@@

The study was carried out in a 10-ha permanent plot established in 1990 by the French Institute of Pondicherry of an undisturbed, old-growth wet evergreen Diptrocarp forest [@pascal_structure_1996]. The plot is located near Uppangala village ("Uppangala plot") in the Kadamakal Reserve Forest (Kodagu District, Karnataka State, India) in India's Western Ghats biodiversity hotspot. Climate is characterized by both heavy rainfall and very dry seasons. Annual precipitation averages 5108 mm per year with 90% between June and October during the monsoon, while the dry season (monthly rainfall $\leq 45$ mm) lasts on average 3.8 months [@pelissier_twenty_2011]. Mean annual temperature in the station of Sampaji (15 km from Uppangala) is about 27$^\circ C$, with a mean minimum about 25$^\circ C$ during July rainfall pick and a maximum mean temperature of 29$^\circ C$ in April [@pascal_structure_1996]. Climate and geology of the 10-ha plot can be considered as homogeneous. General topography of the site shows a regular east-west alternation between thalwegs and ridges separated by relatively steep slopes, so that topography can be assumed as a major source of environmental variation within the forest plot [@gimaret-carpentier_sampling_1998]. We hypothesized that topographic variation should influence community variation depending on tree ecological strategies [@condit_spatial_2000].

## Wood and leaf sampling

Since 1989, all trees $\geq 30$ cm girth at breast height (gbh) have been monitored in an area progressively extended to 10 ha in 2014. All trees were identified, tagged, mapped, and provided with permanent dendrometer bands censused every 2-5 years [@pelissier_twenty_2011]. 111 different species were recorded. We collected trait data on 89 over 111 (80.1%) species, which represented 99.0% of the tree individuals present in the Uppangala plot. With the 32 most abundant species, we reached  more than 80% of Uppangala tree individuals. We sampled five individuals per species for these species, and sampled one to five individuals of the less abundant and rare species (in total 267 individuals were sampled). For each individual we collected 5 mature shadow leaves and one branch sample with a diameter close to 1 cm following recommendation from @perez-harguindeguy_new_2013. Individuals and leaves were selected in consistent light conditions, development stage and sanitary state. Dawkins crown exposure index [@dawkins_management_1958] was assessed for each tree sampled. Leaf and branch samples were kept in humidified ziplock filled with enriched CO2 air inside a black bag untill their measurement in the following 24 hours. For some individuals, we were unable to collect an appropriate branch sample (11 missing species over 89 representing 1% of individuals present in the 10 ha plot).

@@

We calculated five functional traits of leaf and wood samples: leaf thickness (LT in $\mu m$ ), leaf area measured from scans (LA in $mm^2$), leaf dry matter content as the leaf dry weight divided by the leaf fresh weight (LDMC in $mg.g^-1$), specific leaf area as the leaf area divided by the leaf dry mass (SLA in $m^2.kg^-1$), and branch wood density  as the the wood sample dry weight divided by its volume (WD in $g.cm^-3$, see table 1). We choose those five traits because they have been widely used, besides there easier measurement, and they have been shown to be good indicator of species abilities to grow, survive and reproduce [@wilson_specific_1999].

Table 1: Functional traits measured. *Traits with their abbreviation, standard unit, and associated trade-off in tree strategy for functionning, survival and reproduction from @baraloto_decoupled_2010.*

```{r 1 Traits}
rm(list = ls()[-which(ls() == "cores")]); invisible(gc())
PFT <- genPFT()
table <- data.frame(
  Trait = c('Leaf thickness', 'Leaf area', 'Leaf dry matter content', 'Specific leaf area', 'Wood density'),
  Abbreviation = c('LT', 'LA', 'LDMC', 'SLA', 'WD'),
  Unit = c('$\\mu m$', '$mm^2$', '$mg.g^-1$', '$m^2.kg^-1$', '$g.cm^-3$'),
  s = c('Leaf defense vs. investment',
        'Leaf resource capture vs.investment',
        'Leaf defense vs. investment',
        'Leaf resource acquisiton vs. defense',
        'Stem transport, structure and defense')
)
names(table)[4] <- 'Trade-off'
kable(table, row.names = F)
rm(table)
```

@@

Because topography was assumed as a major source of environmental variation [@gimaret-carpentier_sampling_1998], we selected 8 topographical variables representing the abiotic environment: elevation, slope, curvature, plan curvature, profile curvature, wetness (function of both slope and elevation representing water flow and accumulation areas), southwesterness (SW, aspect in direction of south-west: $cos(aspect~in~degres + 225)$) derived from the digital elevation model using the RSAGA package [@brenning_statistical_2008], and distance to the nearest potential river computed using ArcGis [@desktop_arcgis_2011]. 
<!-- Digital elevation model was provided by a 2013 LIDAR campaign on Uppangala plots with a resolution of $1*1 m^2$. -->

In order to identify the main sources of environmental variation and to avoid multi-colinearity in subsequent analyses, we performed a principal component analysis (PCA) of the environmental variables. We selected the less correlated variables with the most ecological meaning. Thus we reduced our explanatory variables to 4 variables: curvature representing the water flow paths and soil accumulation areas, wetness highlighting water accumulation areas, southwesterness representing aspect [@bunyan_effect_2015] and slope.

We selected two biotic factors related to local forest structure: canopy height and basal area. Canopy height was calculated from a digital canopy model built with the 2013 LIDAR campaign on Uppangala plots with resolution $1*1 m^2$. Basal area was calculated from 2013 girth inventory of Uppangala plot on stems with a diameter above 10 centimeters at breast height ($DBH>10~cm$).

@@

The coefficient of variation of traits was investigated at three levels ($cv = \frac{\sigma}{\mu}$ with $\sigma$ the standard deviation and $\mu$ the mean): between leaves of the same individual (intra-individual), between trees of the same species (intra-specific for species with more than one individual), and between species (inter-specific). In addition, we performed a principal component analysis (PCA) of the mean value of traits per species, to characterize basic dimensions of ecological strategies. Subsequent community-level analysis were based on trait values averaged per species.

@@

Across the whole 10 hectare plot, $30*30 m^2$ quadrats were defined and represented 110 local communities. @pascal_wet_1988 defined four categories of strata groups: emergent below 39m, canopy tree between 25 and 39m, sub-canopy tree between 15 and 25m, and understorey tree below 15m. Due to topography, sub-canopy tree from Uppangala are often under light exposure and participates to canopy, especially in slopes. We defined two guilds to represent vertical stratification called canopy (C), composed of emergent, canopy and sub-canopy tree, and understorey (U) trees. The 110 local communities where split by those two guilds and communities accounting for only one species were removed from the subsequent analysis.

To investigate local functional composition of canopy and understorey communities we used community weighted means (CWM), community weighted variances (CWV), and coefficients of variation from the nearest neighbour distance (CVNND) [CWM and CWV, sensu @violle_let_2007; CVNND, @jung_intraspecific_2010]. CWM for each trait $t$ was calculated as the mean of species trait values, $t_{i}$, of the $S$ species in each community, with each species, $i$, weighted by its relative abundance, $p_{i}$:
$$CWM = \sum_{i=1}^{S} p_{i} * t_{i} ~ (1)$$
We also calculated the average of trait values based on species presence-absence, to further emphasize the contribution of less abundant and rare species in community trait distribution [@cingolani_filtering_2007]. CWV for each trait $t$ was calculated as the mean distance of trait values, $t_{i}$, to $CWM$, of the $S$ species in each community, with each species, $i$, weighted by its relative
abundance, $p_{i}$:
$$CWV = \sum_{i=1}^{S} p_{i} * (t_{i} - CWM)^2 ~ (2)$$
CVNND for each trait $t$, scaled with its mean and variance, was calculated as nearest neigbor distances (NND) standard deviation, $\sigma _{NND}$, of the $i$ individuals in each community, divided by its mean, $\mu _{NND}$; and NND for one individual $i$ is calculated as the minimal distance $d$ between the $j$ other individuals of the community :
$$CVNND = \frac{\sigma _{NND}}{\mu _{NND}} ~ , ~ ~ ~ NND = [NND_{i} = min(d_{i,j})_{i \neq j}] ~ (3)$$

@@

We used null models to test if observed correlation with Mantel statistic differed significantly from expectations under null hypothesis. $I_{st}$ correlation to environment and space was tested with 999 repetions of the model nul described by @myers_beta-diversity_2013 which randomizes individuals among subplots while preserving the regional abundance of species, and the number of individuals per community. $U st$ and $\tau st$ were tested by the nul model NM1 which randomizes traits among species.

All analysis were conducted using the R 3.3.0  software environment [@r_core_team_r:_2016] with packages 
`ade4` [@dray_ade4_2007], 
`ape` [@paradis_ape:_2004], 
`phylosignal` [@keck_phylosignal:_2015], 
`spacodiR` [@eastman_spacodir:_2013],
and `vegan` [@oksanen_vegan:_2016].

@@

Table 2: Coefficient of intra-individual, intra-specific and inter-specific trait variation. *Traits with their range of values, mean and standard deviation, mean of coefficient of variation for leaves in all individuals (intra-individual), for trees in all species (intra-specific), and coefficient of variation for trees for all species. See table 1 for abbreviation.*

@@

The principal component analysis (PCA) shows WD associated with structural leaf traits orthogonal to the SLA (figure 1). The first PCA axis (PCA1) covary mainly with LT, LA, LDMC and WD; opposing thick big leaves and lightwood to dense leaves and hardwood. PCA1 axis associated WD to leaf structural traits linked to hydraulic conductive efficiency along the wood economic spectrum [WES, @chave_towards_2009]. The PCA2 covaried mainly positively with SLA and negatively with LDMC along a leaf economic spectrum [LES, @wright_worldwide_2004]. The leaf econmic spectrum opposes high SLA leaves for a strong productivity capacity, with high phyotosynthesis compare to carbon investment, and high LDMC leaves to ensure persistence in time and against attacks. PCA also highlight the strong correlation between LDMC and WD (Pearson's correlation test, $r = 0.25$ with $p-value < 0.001$), indicating a coordinated structural investment of species among stem and leaf.

@@

# Indovidual to log
PFT_ind$LA <- log(PFT_ind$LA)
PFT_ind$SLA <- log(PFT_ind$SLA)
PFT_ind$ID <- row.names(PFT_ind)
row.names(PFT_ind) <- PFT_ind$ID
PFT_ind <- PFT_ind[-which(names(PFT_ind) == 'ID')]
# Species aggregation
Species <- genSpecies()
PFT_sp <- aggregate(PFT_ind, by = list(PFT_ind$Sp), FUN = mean, na.rm = T)
row.names(PFT_sp) <- PFT_sp$Group.1
PFT_sp$SpCode <- SpCode[row.names(PFT_sp)]
PFT_sp <- PFT_sp[-c(1:2,4)]
# PCA
pca <- princomp(~ LA + WD + LT + LDMC + SLA, data = PFT_sp, cor = T)
lambda <- pca$sdev * sqrt(nrow(pca$scores))
pca.var = apply(pca$score, 2, var)
PCAscores <- as.data.frame(pca$scores)[1:2]
names(PCAscores) <- c('WES', 'LES')
PCAscores$ID <- row.names(PCAscores)
plot(t(t(pca$scores)/lambda),pch=16, col = 'lightgrey',
      xlab = paste('Axe 1 -', round(pca.var[1]/sum(pca.var)*100, 2), '%'),
     ylab = paste('Axe 2 -', round(pca.var[2]/sum(pca.var)*100, 2), '%'))
par(new=T)
Rot <- t(t(pca$loadings)*lambda)
XLIM <- c(-max(abs(Rot[,1])),max(abs(Rot[,1])))
XLIM <- XLIM+(XLIM*0.2)
plot(Rot,col=4,axes=FALSE,xlim=XLIM,ylim=XLIM,pch="",xlab = "", ylab = "")
lines(XLIM, c(0,0), lty = 2, col = 'firebrick', lwd = 2)
lines(c(0,0), XLIM, lty = 2, col = 'green', lwd = 2)
legend('topright', c('Leaf economic spectrum', 'Wood economic spectrum'),
       text.col = c("green", "firebrick"))

@@

Figure 1: Principal Component Analysis (PCA) performed on species mean trait values. *The bottom right subplot represents the eigen values (grey bars) compare to a broken stick distribution for 5 components (red line). See table 1 for abbreviation.*

>>>>

Understanding the determinants of species occurrence and co-occurrence is a long standing issue in ecology [@diamond_ecology_1975]. Tropical rainforests have especially fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Many theories have been proposed to explain how species coexistence can be maintained in ecological communities . Among them, two have been hotly debated and tested those last years. First niche theory states that the niche differences of species determine wether they can establish and coexist in communities [@weiher_assembly_1995; @lortie_rethinking_2004], while neutral theory shows that species coexistence is possible even when species are functionaly equivalent, under the sole influence of stochasticity in life, death, reproduction and dispersal [@hubbell_unified_2001]. Both theory can provide realistic expectations of taxonomic diversity [@chave_comparing_2002], and some recent approaches have advocated investigating patterns of functional [@westoby_land-plant_2006; @mcgill_rebuilding_2006] and phylogenetic diversity [@mouquet_ecophylogenetics:_2012] to trace the signature of ecological processes related to specific biological attributes and adaptations. 
These approaches have been applied to resolve the conundrum of high tree diversity in tropical forests [@kraft_functional_2010; @baraloto_using_2012; @hardy_phylogenetic_2012], but studies on these ecosystems still remain scarce. We here investigate taxonomic, functional and phylogenetic diversity in a low elevation rainforest of the Western Ghats biodiversity hotspot [@myers_biodiversity_2000], where few studies have assessed the relative influence of neutral and niche-based processes on local forest composition.

The variation in functional traits among species reflect different abilities to grow, survive and reproduce in a given environment [@violle_let_2007]. These variation thus reflect fundamental trade-offs in ecological strategies along environmental gradients [@mark_westoby_plant_2002]. For instance, specific leaf area varies along a soil fertility-shade index and represent a trade-off between resource acquisition by photosynthesis and leaf investment in defense and leaf lifespan [@hodgson_is_2011]. Due to shared ecological strategies, functional traits will covary along basic leaf [@wright_worldwide_2004 but see @osnas_global_2013 and @lloyd_photosynthetically_2013] and wood [@chave_towards_2009] economics spectra. The leaf economics spectrum opposes productive ecological strategy to conservative persistent strategy. The wood economics spectrum distinguish species strategies over three axes spanning the basic CSR strategies of @grime_evidence_1977 : (i) competitive species, wiht high conductive efficiency, (ii) stress tolerant species, resistant to embolism, and (iii) disturbance tolerant species, with high mechanical strength [@chave_towards_2009]. @baraloto_decoupled_2010 highlighted decoupled wood and leaf strategies allowing independant variation of these strategies, but @reich_world-wide_2014 conversely advocated that the two spectra have to integrated in a comprehensive whole plant economics spectrum [@freschet_evidence_2010] opposing fast growing species to more conservative ones. Here we address how varying wood and leaf ecological strategies determine species community assembly in an Indian rainforest.

A major challenge in addressing community assembly is to relate local coexistence to a broader spatial context of species dynamics [@ricklefs_disintegration_2008; @munoz_neutral_2016]. Niche-based dynamics can entail community variation along broad environmental gradients, while local community dynamics can be neutral [@leibold_coexistence_2006]. Local community dynamics can further be influenced by niche differences under the influence of competition [@macarthur_limiting_1967]. In this perspective, @messier_trait_2016 have underlined that the significance of the leaf economics spectrum is basically scale-dependent, which can explain conflicting results across studies performed at different scales. In the same logic, a spatial hierarchy of ecological dynamics can entail a hierarchy of variation in functional traits from broad variation among species along large-scale environmental gradients, to intraspecific response to micro-environmental conditions at very fine scale [@albert_when_2011]. Therefore a multiscale framework is needed to represent the hierarchy of ecological processes from regional to local scale [@lortie_rethinking_2004]. Determining the extent of neutral and niche-based dynamics thus implies decomposing diversity patterns within and among communities [@ackerly_trait-based_2007], and designing appropriate statistical tests to disentangle the influence of local and larger-scale dynamics [@bernard-verdier_community_2012]. In this paper, we will decompose diversity patterns in a large rainforest plot into sources of variation within and between communities [@kraft_disentangling_2011], and investigate the fine-scale and plot-scale dynamics using appropriate null models. For instance, @messier_trait_2016 argued the scale to which traits are covarying in ecological strategy, more especially they challenged the existence of the leaf economics spectrum at community scale.

Apart from scale-dependent dynamics in space within and among communities, ecological guilds can coexist and occupy distinct niches within communities, while the species within guilds are ecologically similar [@wilson_guilds_1999; @vergnon_niches_2009]. The neutral theory has been basically applied in rainforests to adress diversity patterns in the guild of canopy trees [@hubbell_unified_2001], thus ignoring dynamics of understorey trees. However basic ecological constraints drive resource use strategies and biotic interactions in these ecosystems and imply considering functional strategies and relationship among guilds as well. Low elevation rainforests in South India are remarkably heterogeneous in vertical structure and encompass distinctive guilds across the vertical strata [@pascal_wet_1988]. It thus provide an ideal context to analyze guild-dependent dynamics and ecological strategies in tropical rainforests. We will thus address diversity patterns and community assembly at guild-level.

In this paper, we used a trait-based approach to adress within-guild community dynamics within and among local tree assemblages in a tropical rain forest from Western Ghats, India. Although considered as a wet evergreen forest, the forest undergoes a long dry season and an intense rainy season under the influence of monsoon. In this context, we expect water stress and competition for resource to be the key drivers of forest dynamics. We thus adressed two main questions:

1. Do variations and covariations in functional traits reflect distinct ecological strategies within and among species ?
    i. How functional traits vary among species ?
    ii. Do covariations in functional traits reflect axes of ecological strategy ?
2. Does functional composition depends on potential local abiotic and biotic drivers of community dynamics ?
    i. Do local functional composition ($\alpha$ scale) reflect the influence of abiotic and biotic drivers of community dynamics ?
    ii. Do functional composition variation between communities ($\beta$ scale) reflect the changing of abiotic and biotic conditions ?

@@

The study was carried out in a 10-ha permanent plot established in 1990 by the French Institute of Pondicherry in an undisturbed, old-growth wet evergreen Dipterocarp forest [@pascal_structure_1996]. The plot is located near Uppangala village, and thus refers to as "Uppangala plot", in the Kadamakal Reserve Forest (Kodagu District, Karnataka State, India). Annual precipitation averages 5108 mm per year but is highly variable in time with heavy rainfall from June to October (90% of the rainfall during the monsoon) and a severe dry season (monthly rainfall $\leq 45$ mm) lasting 3.8 months [@pelissier_twenty_2011]. Mean annual temperature in the station of Sampaji (15 km from Uppangala) is about 27$^\circ C$, with a mean minimum about 25$^\circ C$ during July rainfall pick and a maximum mean temperature of 29$^\circ C$ in April [@pascal_structure_1996]. Climate and pedology of the 10-ha plot can be considered as homogeneous. The topography of the site is highly heterogeneous with a regular east-west alternation between thalwegs and ridges separated by steep slopes, so that topography can be assumed as a major source of environmental variation within the forest plot [@gimaret-carpentier_sampling_1998]. We hypothesized that topographic variation should influence community variation depending on tree ecological strategies [@condit_spatial_2000].

## Floristic and trait data

Since 1989, all trees $\geq 30$ cm in girth at breast height (gbh) have been monitored in an area progressively extended to a contiguous 10-ha plot up to 2014. <!-- Need to add Cedric Vega in Acknowledgments for that. --> All trees were identified, tagged and mapped over the plot [@pelissier_twenty_2011]. 111 different species were recorded. We collected trait data on 89 over 111 (80.1%) species, which represented 99.0% of the tree individuals present in the Uppangala plot. Note that the 32 most abundant species, represent  more than 80% of Uppangala tree individuals. We sampled five individuals per species for these 32 species, and sampled one to five individuals from the remaining species (in total 267 individuals were sampled). For each individual we collected 5 mature shadow leaves and one branch sample with a diameter close to 1 cm following recommendation from @perez-harguindeguy_new_2013. Individuals and leaves were selected in consistent light conditions, development stage and sanitary state. Dawkins crown exposure index [@dawkins_management_1958] was recorder for each tree sampled. Leaf and branch samples were kept in humidified ziplock filled with enriched $CO_{2}$ air inside a black bag untill their measurement in the following 24 hours. For some individuals, we were unable to collect an appropriate branch sample (11 missing species over 89 representing 1% of individuals present in the 10 ha plot).

@@

We calculated five functional traits of leaf and wood samples: leaf thickness (LT in $\mu m$ ), leaf area measured from scans (LA in $mm^2$), leaf dry matter content as the leaf dry weight divided by the leaf fresh weight (LDMC in $mg.g^-1$), specific leaf area as the leaf area divided by the leaf dry mass (SLA in $m^2.kg^-1$), and branch wood density  as the the wood sample dry weight divided by its volume (WD in $g.cm^-3$, see table 1). We choose those five traits because they have been widely used, in addition of there easier measurement, and they have been shown to be good indicator of species abilities to grow, survive and reproduce [@wilson_specific_1999].

Table 1: Functional traits measured. *Traits with their abbreviation, standard unit, and associated species strategy for functionning, survival and reproduction from @baraloto_decoupled_2010.*

```{r 1 Traits}
rm(list = ls()[-which(ls() == "cores")]); invisible(gc())
PFT <- genPFT()
table <- data.frame(
  Trait = c('Leaf thickness', 'Leaf area', 'Leaf dry matter content', 'Specific leaf area', 'Wood density'),
  Abbreviation = c('LT', 'LA', 'LDMC', 'SLA', 'WD'),
  Unit = c('$\\mu m$', '$mm^2$', '$mg.g^-1$', '$m^2.kg^-1$', '$g.cm^-3$'),
  Strategy = c('Leaf defense vs. investment',
        'Leaf resource capture vs.investment',
        'Leaf defense vs. investment',
        'Leaf resource acquisiton vs. defense',
        'Stem transport, structure and defense')
)
kable(table, row.names = F)
rm(table)
```

@@

Because topography was assumed as a major source of environmental variation [@gimaret-carpentier_sampling_1998], we selected 8 topographical variables representing the abiotic environment: elevation, slope, curvature, plan curvature, profile curvature, wetness (function of both slope and elevation representing water flow and accumulation areas), southwesterness (SW, aspect in direction of south-west: $cos(aspect~in~degres + 225)$) derived from elevation data available within the plot using the RSAGA package [@brenning_statistical_2008], and distance to the nearest potential river computed using ArcGis [@desktop_arcgis_2011].

In order to identify the main sources of environmental variation and to avoid multi-colinearity in subsequent analyses, we performed a principal component analysis (PCA) of the environmental variables (see supplementary materials S1). We selected the less correlated variables having the most ecological meaning. Thus we reduced our explanatory variables to 4 variables: curvature representing the water flow paths and soil accumulation areas, wetness highlighting water accumulation areas, southwesterness representing aspect [@bunyan_effect_2015] and slope, which may have both a strong effect on forest dynamics and on mutrient/water availability **ref**.

We selected two biotic factors related to local forest structure: canopy height and basal area. Canopy height, i.e. the maximum local height of the canopy, was calculated from a digital canopy model built from a LiDAR (Light Detection And Ranging) campaign done in 2013 over the Uppangala plot with a resolution of $1*1 m^2$. Basal area was calculated from 2013 and 2014 girth inventories of Uppangala plot on stems with a diameter above 30 centimeters at breast height ($DBH>30~cm$).

@@

To quantify trait variation whithin and between individuals and between species we calculated the coefficient of variation (ratio of the standard deviation to the mean) of trait values between leaves of the same individual (intra-individual), between trees of the same species (intra-specific for species with more than one individual), and between species (inter-specific). In addition, to assess the covariation of trait within species, and thus potential axes of strategy, we performed a principal component analysis (PCA) of the mean value of traits per species. Subsequent community-level analysis were based on trait values averaged per species.

@@

Across the whole 10 hectare plot, $30*30 m^2$ quadrats were defined and represented 110 local communities. @pascal_wet_1988 defined four categories of strata groups: emergent below 39m, canopy tree between 25 and 39m, sub-canopy tree between 15 and 25m, and understorey tree below 15m. Due to topography, sub-canopy tree from Uppangala are often under light exposure and participates to canopy, especially in slopes. We defined two guilds to represent vertical stratification called canopy (C), composed of emergent, canopy and sub-canopy tree, and understorey (U) trees. The 110 local communities where split by those two guilds.

To characterize the local functional composition of canopy and understorey communities we used the community weighted mean (CWM), the community weighted variance (CWV), and the coefficient of variation from the nearest neighbour distance (CVNND) [CWM and CWV, sensu @violle_let_2007; CVNND, @jung_intraspecific_2010]. CWM for each trait $t$ was calculated as the mean of species trait values, $t_{i}$, of the $i$ species in the quadrat $j$, with each species, $i$, weighted by its relative abundance, $p_{i}$:
$$CWM_{j} = \sum_{i=1}^{S} p_{i,j} * t_{i,j} ~ (1)$$
We also calculated the average of trait values based on species presence-absence, to further emphasize the contribution of less abundant and rare species in community trait composition [@cingolani_filtering_2007]. CWV for each trait $t$ was calculated as the mean distance of trait values, $t_{i}$, to $CWM$, of the $S$ species in the quadrat $j$, with each species, $i$, weighted by its relative
abundance, $p_{i}$:
$$CWV_{j} = \sum_{i=1}^{S} p_{i,j} * (t_{i,j} - CWM_{j})^2 ~ (2)$$
CVNND for each trait $t$, was calculated as the standard deviation of the nearest neighbor distances (NND), $\sigma _{NND}$, of the $i$ individuals in each community, divided by the mean NND, $\mu _{NND}$; and NND for one individual $i$ is calculated as the minimal distance $d$ between the $j$ other individuals of the community :
$$CVNND = \frac{\sigma _{NND}}{\mu _{NND}} ~ , ~ ~ ~ NND = {NND_{i} = min(d_{i,j})_{i \neq j}} ~ (3)$$

@@

Variation in composition between community pairs with those metrics was translated as a taxonomic, functional and phylogenetic distance, and compared to the corresponding environmental and spatial distances. Environmental distances were calculated as euclidean distances between communities on the plan composed by the two first axes of environmental variables PCA (see supplementary material S1). Correlations with environment and space were assessed with two Mantel tests successively with taxonomic and functional, community composition. The firt Mantel test consisted in assessing the correlation between each metric distance matrix and the environmental distance matrix (E) and the second test, a partial Mantel test, assessed the correlation between each metric distance matrix and the spatial distance matrix knowing environmental distance matrix (E|S). This second test thus test the remaining spatial dependency in community composition once the influence of environmental variation is removed.

We used null models to test if the observed correlations differed significantly from expectations under null hypothesis. $I_{st}$ correlations with environment and space were tested with 999 repetions of the nul model described by @myers_beta-diversity_2013 which randomizes individuals among quadrat while preserving the regional abundance of species, and the number of individuals per community. $U st$ and $\tau st$ were tested by the nul model NM1 which randomizes traits among species.

All analyses were conducted using the R 3.3.0  software environment [@r_core_team_r:_2016] with packages 
`ade4` [@dray_ade4_2007], 
`ape` [@paradis_ape:_2004], 
`phylosignal` [@keck_phylosignal:_2015], 
`spacodiR` [@eastman_spacodir:_2013],
and `vegan` [@oksanen_vegan:_2016].

@@

Table 2: Coefficient of variation of intra-individual, intra-specific and inter-specific trait values. *Traits with their range of values, mean and standard deviation, mean of coefficient of variation for leaves in all individuals (intra-individual), for trees in all species (intra-specific), and coefficient of variation for trees for all species. See table 1 for abbreviation.*

@@

The principal component analysis (PCA) revealed a first axis along which both wood density and structural leaf traits (LT, LDMC) correlate, thus opposing species with thick, big leaves and lightwood to species with dense leaves and hardwood. Interestingly, wood density was orthogonal to SLA in the plan 1-2 of the PCA indicating independence between these two variables. The second axis of the PCA2 covaried mainly positively with SLA and negatively with LDMC. PCA also highlight the strong correlation between LDMC and WD (Pearson's correlation test, $r = 0.25$ with $p-value < 0.001$).

@@

# Indovidual to log
PFT_ind$LA <- log(PFT_ind$LA)
PFT_ind$SLA <- log(PFT_ind$SLA)
PFT_ind$ID <- row.names(PFT_ind)
row.names(PFT_ind) <- PFT_ind$ID
PFT_ind <- PFT_ind[-which(names(PFT_ind) == 'ID')]
# Species aggregation
Species <- genSpecies()
PFT_sp <- aggregate(PFT_ind, by = list(PFT_ind$Sp), FUN = mean, na.rm = T)
row.names(PFT_sp) <- PFT_sp$Group.1
PFT_sp$SpCode <- SpCode[row.names(PFT_sp)]
PFT_sp <- PFT_sp[-c(1:2,4)]
# PCA
pca <- princomp(~ LA + WD + LT + LDMC + SLA, data = PFT_sp, cor = T)
lambda <- pca$sdev * sqrt(nrow(pca$scores))
pca.var = apply(pca$score, 2, var)
PCAscores <- as.data.frame(pca$scores)[1:2]
names(PCAscores) <- c('WES', 'LES')
PCAscores$ID <- row.names(PCAscores)
plot(t(t(pca$scores)/lambda),pch=16, col = 'lightgrey',
      xlab = paste('Axe 1 -', round(pca.var[1]/sum(pca.var)*100, 2), '%'),
     ylab = paste('Axe 2 -', round(pca.var[2]/sum(pca.var)*100, 2), '%'))
par(new=T)
Rot <- t(t(pca$loadings)*lambda)
XLIM <- c(-max(abs(Rot[,1])),max(abs(Rot[,1])))
XLIM <- XLIM+(XLIM*0.2)
plot(Rot,col=4,axes=FALSE,xlim=XLIM,ylim=XLIM,pch="",xlab = "", ylab = "")
lines(XLIM, c(0,0), lty = 2, col = 'firebrick', lwd = 2)
lines(c(0,0), XLIM, lty = 2, col = 'green', lwd = 2)
# legend('topright', c('Leaf economics spectrum', 'Wood economics spectrum'),
       # text.col = c("green", "firebrick"))

@@

Figure 1: Principal Component Analysis (PCA) performed on species mean trait values. *The bottom right subplot represents the eigen values (grey bars) compare to a broken stick distribution for five components (red line). See table 1 for abbreviation.*
